---
title: "Data Story"
author: "Julia Sheriff"
date: "9/22/2018"
output:
  pdf_document: default
  html_document: default
---

## **INTRODUCTION**

   Health insurance companies must create plans that effectively ensure their clients and maximize profits. Because health charges vary from person to person, it is difficult for insurance companies to design insurance plans which collectively maximize profits. Collectively, the insurance company must charge clients more than the cost of covered health charges. We can therefore use individual health charges as an estimate for insurance charges.   
   
   We can begin to estimate individual charges by comparing charges between different clusters of individuals (sex, region, number of children, bmi, region, age).   
   
   By understanding the relationship between charges and these variables, insurance companies can do the following:   
* Predict their charges as their population changes over time.    
* Examine how to provide reimbursement for health services which could make their population less costly.   
* Determine the most locations of the most profitable populations and how to increase clients from that area. Example: Choosing a location for an HMO.

*** 

## **OVERVIEW OF THE DATASET**

**Health Variables:**

The dataset is available at <https://www.kaggle.com/mirichoi0218/insurance/home>.

|Variable|Description|
|:-------|:------------------------------------------------------------------------------|
|Age| individual's age in years
|Sex| insurance contractor gender: female, male 
|BMI| Body mass index: weight in kg / heght in m^2
|BMI_factor| Categories of BMI values: underweight, healthy weight, overweight, obese
|Children| Number of children covered by health insurance, Number of dependents
|Smoker| Smoker or Non-smoker
|Region| Beneficiary's US residental area: northeast, southeast, northwest, southwest
|Charges| Individual medical costs billed by health insurance

```{r}
health_charges <- read.csv("capstone_data.csv", header = TRUE)
head(health_charges)
str(health_charges)
```
***

## **CONSIDERATIONS**   

*While we have data on seven variables in our observations, there are other factors which could impact health charges.*    
* income of individual   
* education level    
* employment status   
* location: urban, suburban, rural   
* chronic health conditions   
* muscle / fat ratio (in addition to BMI which just compares weight to height)   

*There are also other factors that would be useful in interpreting the charges themselves:*
* breakdown of charges   
* charges from previous years

***

## **DATA CLEANING** 

I created a new variable, BMI_factor, which treats the numerical BMI variable as a factor variable of four categories: underweight, healthy weight, overweight, obese. These are standard categories used by agencies such as the CDC. I created this variable to give a general meaning to the numerical variable, bmi.   
```{r}
library (dplyr)
health_charges <- health_charges %>% 
  mutate(bmi_factor = if_else ( bmi < 18.5, "underweight",
         if_else ( bmi >= 18.5 & bmi < 25, "healthy_weight", 
         if_else ( bmi >= 25 & bmi < 30, "overweight",
         if_else ( bmi >= 30, "obese", NA_character_)))))

health_charges$bmi_factor <- factor(health_charges$bmi_factor, 
                      levels = c("underweight", "healthy_weight", "overweight", "obese"),
                      ordered = TRUE)

health_charges <- health_charges[ , c(1:3, 8, 4:7)]

health_charges_clean <- health_charges

head(health_charges_clean)

```

I assessed the data for missing values and nonsensical outliers, and the data was clean. The data was tidy because each row represents and observation and each column represents a variable.    
```{r, eval=F, echo=T}
summary(health_charges == "")
summary(is.na.data.frame(health_charges)) 

unique(health_charges[,1])
unique(health_charges[,2])
unique(health_charges[,3])
unique(health_charges[,4])
unique(health_charges[,5])
unique(health_charges[,6])
unique(health_charges[,7])
unique(health_charges[,8])

head(sort(health_charges$bmi), n=25)
tail(sort(health_charges$bmi), n=25)
head(sort(health_charges$charges), n=25)
head(sort(health_charges$charges), n=25)
```
***

## **INITIAL FINDINGS**
### **UNIVARIATE ANALYSIS**
**AGE**  
* Disporportionately high number of 18-19 ages;  
* Otherwise, even age distribution.   
**SEXES**  
* Even distribution    
**BMI and BMI_FACTOR**  
* Normal distribution  
* The mean of the data is approximately at the border of overweight and obese.  
* The number of obese observations is approximately equal to the sum of the non-obese observations.
**CHILDREN**  
* The data is skewed right.     
**SMOKER**  
* The ratio of non-smokers to smokers is approximately 4 : 1     
**REGION**   
* All regions except southeast had between 324-325 observations.
* Perhaps cluster sampling was used for data collection.     
**CHARGES**   
* SHAPIRO.TEST   
  + HO: Charges frequency follows a normal distribution.   
  + HA: Charges frequency does not follow a normal distribution.   
  + RESULTS:   
    - P-Value: < 2.2e-16 < .05   
    - Reject HO.   
    - Evidence supports the claim that charges frequency does not follow a normal distribution.

```{r}
shapiro.test(health_charges_clean$charges)
```

### **MULTIVARIATE ANALYSIS**  
*Relationships between multiple variables with anecdotal notes*
```{r include=FALSE}
read.csv("health_charges_clean.csv")
```


```{r echo=FALSE, fig.height=6, fig.width=8}
library(ggplot2)
ggplot(health_charges_clean, 
       aes(x = bmi, y = charges, color = bmi_factor, alpha = .005 ))+
  geom_point() +
  geom_jitter() +
  geom_smooth (method = "loess", color = "black")
```

### **Effect of BMI on charges**  
* Charges increase with higher BMIs.  
* There is a positive linear correlation between charges and bmi less than 35.  
* There is no meaningful correlation between charges and bmi above 35.
\newpage
```{r echo=FALSE, fig.height= 6, fig.width=6}

health_charges_clean$children <- as.factor(health_charges_clean$children)
ggplot(health_charges_clean, aes(x = children, y = charges, color = sex)) +
  geom_bar(stat = "identity", aes(color = sex, fill = sex), 
           width = .7, position = "dodge")
```

### **Effect of children on charges, considering sex**  
* Charges decrease with higher numbers of children.  
* Women do not have higher health charges than men in regard to the number of children.
\newpage
```{r echo=FALSE, fig.height=6, fig.width=8}
ggplot(health_charges_clean, aes(x = age, y = charges, color = bmi_factor), alpha = .02, size = .02) +
  geom_point(aes(color = bmi_factor, fill = bmi_factor))+
  facet_grid( . ~ smoker)+
  geom_smooth(se = FALSE, method = "loess", weight = .005, color = "black", alpha = .02 )
```

### **Timeseries of charges, considering BMI and smoking**  
* Smokers have higher charges than non-smokers.  
* Smokers see a strong positive correlation between a higher BMI and charges.  
* Obese smokers have higher charges than most non-smokers of all BMIs.
\newpage
```{r echo=FALSE, fig.height=6, fig.width=8}
ggplot(health_charges_clean, aes(x = region, y = charges, color = bmi_factor))+
  geom_bar(stat = "identity", position = "dodge", 
           aes(color = bmi_factor, fill = bmi_factor), width = .7)
```

### **Region's effect on charges, considering BMI**     
* There were no underweight observations in the southeast region.   
* BMI is a stronger indicator for charges in the south than in the north.

### *EXPLORATION OF OUTLIER REMOVAL*

```{r}

library(ggplot2)

health_charges_clean <- read.csv("health_charges_clean.csv", header=TRUE)

swdf <- health_charges_clean[which(health_charges_clean$region == "southwest") , ]

swmoddf <- swdf
swmod <- swmoddf$charges
qnt <- quantile(swmod, probs=c(.25, .75), na.rm = T)
H <- 1.5 * IQR(swmod, na.rm = T)
swmod[swmod < (qnt[1] - H)] <- (qnt[1] - H)
swmod[swmod > (qnt[2] + H)] <- (qnt[2] + H)
swmoddf$charges <- swmod

min(swdf$charges)
min(swmoddf$charges)
max(swdf$charges)
max(swmoddf$charges)

ggplot(swdf, aes(x = region, y = charges)) +
  geom_boxplot( alpha = .2) +
  ggtitle("SW Charges, outliers intact")

ggplot(swmoddf, aes(x = region, y = charges)) +
         geom_boxplot( alpha = .2) +
         ggtitle("SW Charges, outliers removed")


nwdf <- health_charges_clean[which(health_charges_clean$region == "northwest") , ]

nwmoddf <- nwdf
nwmod <- nwmoddf$charges
qnt <- quantile(nwmod, probs=c(.25, .75), na.rm = T)
H <- 1.5 * IQR(nwmod, na.rm = T)
nwmod[nwmod < (qnt[1] - H)] <- (qnt[1] - H)
nwmod[nwmod > (qnt[2] + H)] <- (qnt[2] + H)
nwmoddf$charges <- nwmod

min(nwdf$charges)
min(nwmoddf$charges)
max(nwdf$charges)
max(nwmoddf$charges)

ggplot(nwdf, aes(x = region, y = charges)) +
  geom_boxplot( alpha = .2) +
  ggtitle("NW charges, outliers intact")
 
ggplot(nwmoddf, aes(x = region, y = charges)) +
         geom_boxplot(alpha = .2) +
         ggtitle("NW charges, outliers removed") 

```

### **BOXPLOTS SHOWING OUTLIERS
```{r}

ggplot(health_charges_clean, aes(x = region, y = charges))+
  geom_boxplot( alpha = .2)

agef <- as.factor(health_charges_clean$age)
ggplot(health_charges_clean, aes(x = agef, y = charges))+ 
geom_boxplot(color = "black", size = 1, alpha = .2) +
geom_jitter(color = "green", size = 1, alpha = .2)


ggplot(health_charges_clean, aes(x = sex, y = charges))+ 
geom_boxplot(color = "black", size = 1, alpha = .2) +
geom_jitter(color = "green", size = 1, alpha = .2) 


ggplot(health_charges_clean, aes(x = bmi_factor, y = charges))+ 
geom_boxplot(color = "black", size = 1, alpha = .2) +
geom_jitter(color = "green", size = 1, alpha = .2)

childrenf <- as.factor(health_charges_clean$children)
ggplot(health_charges_clean, aes(x = children, y = charges))+ 
geom_boxplot(color = "black", alpha = .2) +
geom_jitter(color = "green", size = .5, alpha = .2) 


ggplot(health_charges_clean, aes(x = smoker, y = charges))+ 
geom_boxplot(color = "black", alpha = .2) +
geom_jitter(color = "green", size = .5, alpha = .2) 


ggplot(health_charges_clean, aes(x = region, y = charges))+ 
geom_boxplot(color = "black", alpha = .2) +
geom_jitter(color = "green", size = .5, alpha = .2) 



ggplot(health_charges_clean, aes(x = region, y = charges))+ 
geom_boxplot(color = "black", alpha = .2) +
geom_jitter(color = "green", size = .5, alpha = .2) 

```

## **APPROACH**   
* The approach has not changed. I will apply statistical analysis and machine learning to the data to study the relationship of different variables on charges.





