---
title: "Statistical Analysis of Health Charges"
author: "Julia Sheriff"
date: "9/12/2018"
output: pdf_document
---
***

## **An Overview of the Dataset**

**Health Variables:**

|Variable|Description|
|:-------|:------------------------------------------------------------------------------|
|Age| individual's age in years
|Sex| insurance contractor gender: female, male 
|BMI| Body mass index: weight in kg / heght in m^2
|BMI_factor| Categories of BMI values: underweight, healthy weight, overweight, obese
|Children| Number of children covered by health insurance, Number of dependents
|Smoker| Smoker or Non-smoker
|Region| Beneficiary's US residental area: northeast, southeast, northwest, southwest
|Charges| Individual medical costs billed by health insurance

```{r}

health_charges_clean <- read.csv("health_charges_clean.csv", header=TRUE)

head(health_charges_clean)

```
\newpage

# **Single Variable Analysis**
## An overview of each variable with anecdotal notes  

```{r fig.width = 5, fig.height = 5}
library(ggplot2)

ggplot(health_charges_clean, aes(age))+
  geom_histogram(binwidth = 1)+
  coord_cartesian(xlim = c(18, 64))+
  theme_bw()
```

### **Age**  
* Disporportionately high number of 18-19 ages;  
* Otherwise, even age distribution.  
\newpage
```{r fig.width = 4, fig.height = 4}
ggplot(health_charges_clean, aes(sex))+
  geom_bar()
```

### **Sexes**  
* Even distribution
\newpage
```{r fig.width = 6, fig.height = 6}
ggplot(health_charges_clean, aes(bmi)) +
  geom_histogram(binwidth = 1) +
  coord_cartesian(xlim = c(15, 54))
```

### **BMI**  
* Normal distribution  
* The mean of the data is approximately at the border of overweight and obese.  
* The number of obese observations is approximately equal to the sum of the non-obese observations.
\newpage
```{r fig.width = 4, fig.height = 4}
health_charges_clean$bmi_factor <- factor(health_charges_clean$bmi_factor, 
      levels = c("underweight", "healthy_weight", "overweight", "obese"),
      ordered = TRUE)
ggplot(health_charges_clean, aes(bmi_factor)) +
  geom_bar()
```

### **BMI_factor**  
* More observations for higher BMI categories
\newpage
```{r fig.width = 5, fig.height = 5}
ggplot(health_charges_clean, aes(children))+
  geom_bar()
```

### **Children**  
* The data is skewed right.
\newpage
```{r fig.width = 4, fig.height = 4}
ggplot(health_charges_clean, aes(smoker))+
  geom_bar()
```

### **Smoker**  
* The ratio of non-smokers to smokers is approximately 4 : 1 
\newpage
```{r fig.width = 4, fig.height = 4}
ggplot(health_charges_clean, aes(region))+
  geom_bar()
```

### **Region**   
* All regions except southeast had between 324-325 observations.
* Perhaps cluster sampling was used for data collection.
\newpage
```{r fig.width = 6, fig.height = 6}
ggplot(health_charges_clean, aes(charges)) +
  geom_area(stat = "bin")
```

### **Charges**  
* Skewed right 
\newpage

# **Multivariable analysis**  
## Relationships between multiple variables with anecdotal notes

```{r fig.width = 8, fig.height = 5}
ggplot(health_charges_clean, 
       aes(x = bmi, y = charges, color = bmi_factor, alpha = .005 ))+
  geom_point() +
  geom_jitter() +
  geom_smooth (method = "loess", color = "black")
```

### **Effect of BMI on charges**  
* Charges increase with higher BMIs.  
* There is a positive linear correlation between charges and bmi less than 35.  
* There is no meaningful correlation between charges and bmi above 35.
\newpage
```{r fig.width = 6, fig.height = 4}
health_charges_clean$children <- as.factor(health_charges_clean$children)

ggplot(health_charges_clean, aes(x = children, y = charges, color = sex)) +
  geom_bar(stat = "identity", aes(color = sex, fill = sex), 
           width = .7, position = "dodge")
```

### **Effect of children on charges, considering sex**  
* Charges decrease with higher numbers of children.  
* Women do not have higher health charges than men in regard to the number of children.
\newpage
```{r fig.width = 8, fig.height = 6}
ggplot(health_charges_clean, aes(x = age, y = charges, color = bmi_factor), alpha = .02, size = .02) +
  geom_point(aes(color = bmi_factor, fill = bmi_factor))+
  facet_grid( . ~ smoker)+
  geom_smooth(se = FALSE, method = "loess", weight = .005, color = "black", alpha = .02 )
```

### **Timeseries of charges, considering BMI and smoking**  
* Smokers have higher charges than non-smokers.  
* Smokers see a strong positive correlation between a higher BMI and charges.  
* Obese smokers have higher charges than most non-smokers of all BMIs.
\newpage
```{r fig.width = 8, fig.height = 6}
ggplot(health_charges_clean, aes(x = region, y = charges, color = bmi_factor))+
  geom_bar(stat = "identity", position = "dodge", 
           aes(color = bmi_factor, fill = bmi_factor), width = .7)
```

### **Region's effect on charges, considering BMI**  
* There were no underweight observations in the southeast region.
* BMI is a stronger indicator for charges in the south than in the north.


