---
title: "Statistical Analysis of Health Charges"
author: "Julia Sheriff"
date: "9/12/2018"
output: pdf_document
---
***

## **An Overview of the Dataset**

**Health Variables:**

|Variable|Description|
|:-------|:------------------------------------------------------------------------------|
|Age| individual's age in years
|Sex| insurance contractor gender: female, male 
|BMI| Body mass index: weight in kg / heght in m^2
|BMI_factor| Categories of BMI values: underweight, healthy weight, overweight, obese
|Children| Number of children covered by health insurance, Number of dependents
|Smoker| Smoker or Non-smoker
|Region| Beneficiary's US residental area: northeast, southeast, northwest, southwest
|Charges| Individual medical costs billed by health insurance

```{r}

health_charges_clean <- read.csv("health_charges_clean.csv", header=TRUE)

head(health_charges_clean)

```
\newpage

# **Single Variable Analysis**
## An overview of each variable with anecdotal notes  

```{r fig.width = 5, fig.height = 5}
library(ggplot2)
library(dplyr)
library(Hmisc)

ggplot(health_charges_clean, aes(age))+
  geom_histogram(binwidth = 1)+
  coord_cartesian(xlim = c(18, 64))+
  theme_bw()
```

### **Age**  
* Disporportionately high number of 18-19 ages;  
* Otherwise, even age distribution.  
\newpage
```{r fig.width = 4, fig.height = 4}
ggplot(health_charges_clean, aes(sex))+
  geom_bar()
```

### **Sexes**  
* Even distribution
\newpage
```{r fig.width = 6, fig.height = 6}
ggplot(health_charges_clean, aes(bmi)) +
  geom_histogram(binwidth = 1) +
  coord_cartesian(xlim = c(15, 54))
```

### **BMI**  
* Normal distribution  
* The mean of the data is approximately at the border of overweight and obese.  
* The number of obese observations is approximately equal to the sum of the non-obese observations.
\newpage
```{r fig.width = 4, fig.height = 4}
health_charges_clean$bmi_factor <- factor(health_charges_clean$bmi_factor, 
      levels = c("underweight", "healthy_weight", "overweight", "obese"),
      ordered = TRUE)
ggplot(health_charges_clean, aes(bmi_factor)) +
  geom_bar()
```

### **BMI_factor**  
* More observations for higher BMI categories
\newpage
```{r fig.width = 5, fig.height = 5}
ggplot(health_charges_clean, aes(children))+
  geom_bar()
```

### **Children**  
* The data is skewed right.
\newpage
```{r fig.width = 4, fig.height = 4}
ggplot(health_charges_clean, aes(smoker))+
  geom_bar()
```

### **Smoker**  
* The ratio of non-smokers to smokers is approximately 4 : 1 
\newpage
```{r fig.width = 4, fig.height = 4}
ggplot(health_charges_clean, aes(region))+
  geom_bar()
```

### **Region**   
* All regions except southeast had between 324-325 observations.
* Perhaps cluster sampling was used for data collection.
\newpage
```{r fig.width = 6, fig.height = 6}
ggplot(health_charges_clean, aes(charges)) +
  geom_area(stat = "bin")
```

### **Charges**  
* Skewed right 
\newpage

# **Multivariable analysis**  
## Relationships between multiple variables with anecdotal notes

```{r fig.width = 8, fig.height = 5}
ggplot(health_charges_clean, 
       aes(x = bmi, y = charges, color = bmi_factor, alpha = .005 ))+
  geom_point() +
  geom_jitter() +
  geom_smooth (method = "loess", color = "black")
```

### **Effect of BMI on charges**  
* Charges increase with higher BMIs.  
* There is a positive linear correlation between charges and bmi less than 35.  
* There is no meaningful correlation between charges and bmi above 35.
\newpage
```{r fig.width = 6, fig.height = 4}
health_charges_clean$children <- as.factor(health_charges_clean$children)

ggplot(health_charges_clean, aes(x = children, y = charges, color = sex)) +
  geom_bar(stat = "identity", aes(color = sex, fill = sex), 
           width = .7, position = "dodge")
```

### **Effect of children on charges, considering sex**  
* Charges decrease with higher numbers of children.  
* Women do not have higher health charges than men in regard to the number of children.
\newpage
```{r fig.width = 8, fig.height = 6}
ggplot(health_charges_clean, aes(x = age, y = charges, color = bmi_factor), alpha = .02, size = .02) +
  geom_point(aes(color = bmi_factor, fill = bmi_factor))+
  facet_grid( . ~ smoker)+
  geom_smooth(se = FALSE, method = "loess", weight = .005, color = "black", alpha = .02 )
```

### **Timeseries of charges, considering BMI and smoking**  
* Smokers have higher charges than non-smokers.  
* Smokers see a strong positive correlation between a higher BMI and charges.  
* Obese smokers have higher charges than most non-smokers of all BMIs.
\newpage
```{r fig.width = 8, fig.height = 6}
ggplot(health_charges_clean, aes(x = region, y = charges, color = bmi_factor))+
  geom_bar(stat = "identity", position = "dodge", 
           aes(color = bmi_factor, fill = bmi_factor), width = .7)
```

### **Region's effect on charges, considering BMI**     
* There were no underweight observations in the southeast region.   
* BMI is a stronger indicator for charges in the south than in the north.

\newpage

# **Statistical tests**
**Parametric and non-parametric tests with graphical representations.**

### **ANOVA test, comparing the true mean BMI of adults with different numbers of children**   
* HO:The true mean BMI for adults with different numbers of children is uniform, at a .05 significance level.     
* HA:The true mean BMI for adults with different numbers of children is not uniform, at .05 significance level.     
* RESULT:   
    + P = .883 > .05.       
    + Fail to reject HO.         
    + There is not enough evidence to support that the true mean BMI for adults with different numbers of children is not uniform, at .05 significance level.         
 
```{r}
group_by(health_charges_clean, children) %>%
  summarise(
    count = n(),
    mean = mean(bmi, na.rm = TRUE),
    sd = sd(bmi, na.rm = TRUE)
    )

aov_childrenbmi <- aov(bmi ~ children, data = health_charges_clean)
summary(aov_childrenbmi)

```
\newpage
```{r}
ggplot(health_charges_clean, aes(x=children, y=bmi)) + 
  geom_boxplot(color = "black", alpha = .2) +
  geom_jitter(color = "blue", size = .5, alpha = .2) +
  stat_summary(fun.data = mean_sdl, fun.args = list(mult = 1),        geom = "errorbar", color = "red", width = .2) + 
  stat_summary(fun.data = mean_sdl, fun.args = list(mult = 1), geom = "point", color = "red", size = 2)
```

\newpage

### *Kruskal-Wallis Test, comparing median health charges for adults with different numbers of children*    
* HO: The median health charges between adults with different numbers of children are equal, at a .05 significance level.   
* HA: The median health charges between adults with different numbers of children are unequal, at .05 significance level.   
* RESULT:   
    + P = 1.86e-05 < .05.      
    + Reject HO.   
    + Evidence supports that the median health charges between adults with different numbers of children are unequal. 

```{r}
group_by(health_charges_clean, children) %>%
  summarise(
    count = n(),
    mean = mean(charges, na.rm = TRUE),
    sd = sd(charges, na.rm = TRUE)
    )

kruskal.test(charges ~ children, data = health_charges_clean)

```
\newpage

```{r}
ggplot(health_charges_clean, aes(x=children, y=charges)) + 
  geom_boxplot(color = "black", alpha = .2) +
  geom_jitter(color = "green", size = .5, alpha = .2) +
  stat_summary(fun.data = mean_sdl, fun.args = list(mult = 1), 
    geom = "errorbar", color = "red", width = .2) +
  stat_summary(fun.data = mean_sdl, fun.args = list(mult = 1), 
    geom = "point", color = "red", size = 2)
```

\newpage 

### *Independent T-Test, comparing mean bmi between sexes*   
* HO: Both sexes have the same true mean bmi, at a .05 significance level.   .
* HA: Sexes have a different true mean bmi, at a .05 significance level.    
* RESULTS:    
    + P = .08992 > .05.   
    + Fail to reject HO.     
    + There is not enough evidence to support that sexes have a different true mean bmi, at a .05 significance level. 

```{r}
group_by(health_charges_clean, sex) %>%
  summarise(
    count = n(),
    mean = mean(charges),
    sd = sd(charges)
  )

t.test(bmi ~ sex, data = health_charges_clean)
```
\newpage
```{r}
ggplot(health_charges_clean, aes(x=sex, y=bmi, color =sex)) + 
  geom_jitter(size = 2, alpha = .2) +
  stat_summary(fun.data = mean_sdl, fun.args = list(mult = 1),        geom = "errorbar", color = "black", size = .5, width = .2) + 
  stat_summary(fun.data = mean_sdl, fun.args = list(mult = 1), geom = "point", color = "black", size = 2)
```

\newpage

### *Independent T-Test, comparing mean bmi between smokers and non-smokers*     
* HO: Smokers and non-smokers have the same true mean bmi, at a .05 significance level.   
* HA: Smokers and non-smokers have a different true bmi, at a .05 significance level.   
* RESULTS:    
    + P = 0.8938 > .05.     
    + Failt to reject HO.        
    + There is not enough evidence to support that smokers and non-smokers have a different true bmi, at a .05 significance level.

```{r} 
group_by(health_charges_clean, smoker) %>%
  summarise(
    count = n(),
    mean = mean(charges),
    sd = sd(charges)
  )

  t.test(bmi ~ smoker, data = health_charges_clean)
```
  
\newpage

```{r}
library(ggplot2)
  ggplot(health_charges_clean, aes(x=smoker, y=bmi, color =smoker)) + 
  geom_jitter(size = 2, alpha = .2) +
  stat_summary(fun.data = mean_sdl, fun.args = list(mult = 1),        geom = "errorbar", color = "black", size = .5, width = .2) + 
  stat_summary(fun.data = mean_sdl, fun.args = list(mult = 1), geom = "point", color = "black", size = 2) 
```

\newpage

### *MANN-WHITNEY-WILCOXON TEST, comparing charges between the sexes*       
* HO: The charges of females and males have identical distributions of charges at a .05 significance level.     
* HA: The charges of females and males have different distributions of charges at a .05 significance level.      
* RESULTS:   
    + P=.7287 > .05.     
    + Fail to reject HO.     
    + There is not enough evidence to prove that the charges of females and males have different distributions of charges at a .05 significance level

```{r}

group_by(health_charges_clean, sex) %>%
  summarise(
    count = n(),
    mean = mean(charges),
    sd = sd(charges)
  )

wilcox.test( charges ~ sex, data = health_charges_clean)
```

\newpage

```{r}
ggplot(health_charges_clean, aes(x=sex, y=charges, color =sex)) + 
  geom_jitter(size = 3, alpha = .2) +
  stat_summary(fun.data = mean_sdl, fun.args = list(mult = 1),        geom = "errorbar", color = "black", size = .5, width = .2) + 
  stat_summary(fun.data = mean_sdl, fun.args = list(mult = 1), geom = "point", color = "black", size = 2)
```

\newpage 

### *MANN-WHITNEY-WILCOXON TEST, comparing charges between smokers and non-smokers*       
* HO: The charges of smokers and non-smokers have identical distributions of charges at a .05 significance level.     
* HA: The charges of smokers and non-smokers have different distributions of charges at a .05 significance level.      
* RESULTS:    
    + P < 2.2e-16 < .05.     
    + Reject HO.     
    + Evidence supports that the charges of smokers and non-smokers have different distributions of charges at a .05 significance level

```{r}
library(dplyr)
group_by(health_charges_clean, smoker) %>%
  summarise(
    count = n(),
    mean = mean(charges),
    sd = sd(charges)
  )

wilcox.test( charges ~ smoker, data = health_charges_clean)
```

\newpage

```{r}
ggplot(health_charges_clean, aes(x=smoker, y=charges, color =smoker)) + 
  geom_jitter(size = 4, alpha = .2) +
  stat_summary(fun.data = mean_sdl, fun.args = list(mult = 1),        geom = "errorbar", color = "black", size = .5, width = .2) + 
  stat_summary(fun.data = mean_sdl, fun.args = list(mult = 1), geom = "point", color = "black", size = 2) +
  labs(title = "Distributions of Charges for Smokers and Non-Smokers") +
  scale_fill_manual(name = "Status", labels = c("Nonsmoker", "Smoker"))

```


\newpage

### *PEARSON'S LINEAR REGRESSION, describing the linear relationship between bmi and charges*     
* HO: The true correlation between bmi and charges is equal to 0 at a .05 significance level.   
* HA: The true correlation between bmi and charges is not equal to 0 at a .05 significance level.    
* RESULTS:        
    + P-Value = 2.459e-13 < .05.     
    + Reject HO.   
    + Evidence supports that the true correlation between bmi and charges is not equal to 0 at a .05 significance level.   
    + The true correlation between bmi and charges is .198341, with CI = 0.1463052, 0.2492822.    
    + There is a weak positive correlation between bmi and charges.

```{r}
cor.test(health_charges_clean$bmi, health_charges_clean$charges, method = "pearson")
```

\newpage


```{r}
BMI_Group <- health_charges_clean$bmi_factor
ggplot(health_charges_clean, aes(x = bmi, y = charges))+
  geom_point(size = 2, alpha = .3, aes(color = BMI_Group))+
  geom_smooth(aes(x = bmi, y = charges), method = lm) + 
  labs(title = "Linear Regression of BMI on Charges")
```

\newpage

### *PEARSON'S LINEAR REGRESSION, describing the linear relationship between bmi and charges, subset by bmi_factor      
* HO: The true correlation between bmi and charges is equal to 0 at a .05 significance level.   
* HA: The true correlation between bmi and charges is not equal to 0 at a .05 significance level.
* RESULTS: 
  + Underweight:   
    - p-value = 0.071 > .05, fail to reject HO.   
    - There is not enough evidence to support the claim that the true correlation between bmi and charges is not equal to 0 at a .05 significance level.   
    - 95 percent confidence interval: (-0.03721726,  0.72280204).   
    - Coefficient: 0.4120904   
    - Moderate positive correlation between underweight bmi and charges.   
  + Healthy_weight:  
    - p-value = 0.006103 < .05, reject HO.   
    - Evidence supports the claim that the true correlation between bmi and charges is not equal to 0 at a .05 significance level.   
    - 95 percent confidence interval: (0.05276277, 0.30579513).   
    - Coefficient: 0.1822954   
    - Weak positive correlation between underweight bmi and charges.    
  + Overweight:    
    - p-value = 0.839 > .05, fail to reject HO.    
    - There is not enough evidence to support the claim that the true correlation between bmi and charges is not equal to 0 at a .05 significance level.   
    - 95 percent confidence interval: (-0.11007646,  0.08953425).   
    - Coefficient: -0.01037446   
    - There is a negligiblely weak correlation between overweight bmi and charges. 
  + Obese:    
    - p-value = 0.09527 > .05, fail to reject HO.   
    - There is not enough evidence to support the claim that the true correlation between bmi and charges is not equal to 0 at a .05 significance level.   
    - 95 percent confidence interval: (-0.01099551,  0.13589593).   
    - Coefficient: 0.06279025   
    - There as a negligbly weak positive correlation between overweight bmi and charges.  

```{r}

underweight <- subset(health_charges_clean, bmi < 18.5, select = c(bmi))
ucharges <- subset(health_charges_clean, bmi < 18.5, select = c(charges))
cor.test(underweight[ ,1], ucharges[ ,1], method = "pearson")

healthy_weight <- subset(health_charges_clean, bmi >= 18.5 & bmi < 25, select = c(bmi))
hwcharges <- subset(health_charges_clean, bmi >= 18.5 & bmi < 25, select = c(charges))
cor.test(healthy_weight[ ,1], hwcharges[ ,1], method = "pearson")

overweight <- subset(health_charges_clean, bmi >= 25 & bmi < 30, select = c(bmi))
ovcharges <- subset(health_charges_clean, bmi >= 25 & bmi < 30, select = c(charges))
cor.test(overweight[ ,1], ovcharges[ ,1], method = "pearson")

obese <- subset(health_charges_clean, bmi >= 30, select = c(bmi))
obcharges <- subset(health_charges_clean, bmi >= 30, select = c(charges))
cor.test(obese[ ,1], obcharges[ ,1], method = "pearson")

```

\newpage

```{r}
BMI_Group <- health_charges_clean$bmi_factor
ggplot(health_charges_clean, aes(x = bmi, y = charges, color = BMI_Group))+
  geom_point(size = 1, alpha = .2)+
  geom_smooth(aes(x = bmi, y = charges), method = lm)+
  labs(title = "Linear Regression of BMI Groups on Charges", y = "Charges", x = "BMI ")+
  guides(colorbar = "BMI Groups")
```

\newpage

### *PEARSON'S LINEAR REGRESSION, describing the linear relationship between age and charges*   
* HO: The true correlation between age and charges is equal to 0 at a .05 significance level.   
* HA: The true correlation between age and charges is not equal to 0 at a .05 significance level.    
* RESULTS:        
    + P-Value: < 2.2e-16 < .05.    
    + Reject HO. 
    + Evidence supports the claim that the true correlation between age and charges is not equal to 0 at a .05 significance level.
    + The true correlation between bmi and charges is 0.2990082, with CI = 0.2494139, 0.3470381.    
    + There is a weak positive correlation between bmi and charges.
    
    ```{r}
cor.test(health_charges_clean$age, health_charges_clean$charges, method = "pearson")
```

\newpage

```{r}
ggplot(health_charges_clean, aes(x = age, y = charges))+
  geom_point(size = 3, alpha = .2)+
  geom_smooth(method = lm) +
  ggtitle("Pearson Linear Regression of Age on Charges")
```

\newpage
### *CHI-SQUARED TEST FOR INDEPENDENCE, between bmi group and region       
* HO: Bmi group is independent of region at a .05 significance level.  
* HA: Bmi group is dependent on region at a .05 significance level.  
* RESULTS:   
  + P-Value: 4.015e-09 < .05   
  + Reject HO.   
  + Evidence supports that bmi group is dependent on region at a .05 significance level.    
  
```{r}
chisq.test(health_charges_clean$region, health_charges_clean$bmi_factor)
chisq <- chisq.test(health_charges_clean$region, health_charges_clean$bmi_factor)
chisq$observed
```
\newpage

```{r}

ggplot(health_charges_clean, aes(x = region, y = bmi_factor)) +
  geom_jitter(alpha = .2, size = 8) + 
  ggtitle("Scatterplot of Chi-Squared Distribution between Region and BMI Group") + 
  ylab("BMI Group") +
  xlab("Region")

library(corrplot)
corrplot(chisq$residuals, is.cor = FALSE) 
```



