---
title: "machine learning approach"
author: "Julia Sheriff"
date: "10/28/2018"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

###GENERAL IDEA OF DATASET

```{r}
health_charges_clean <- read.csv("health_charges_clean.csv")
str(health_charges_clean)
head(health_charges_clean)
```

###PREPPING MY VARIABLES
* CHARGES - dependent variable for all predictions
    + continuous
    + CREATE 2 binary vars (above .75 IQR, below .75 IQR)
* Age
    + continuous
    + CREATE 5 binary vars (10 year groups (18-25 counted as one bracket))
* Bmi 
    + continuous
    + categorical (I now have 4 categories, but am adding two more (breaking up obese into 3 categories according to CDC)).
    + CREATE 6 binary vars 
* Sex 
    + binary
* Children 
    + categorical
    + CREATE 5 binary vars 
* Smoker 
    + binary 
* Region 
    + categorical
    + CRETE 4 binary vars 
    
###LINEAR REGRESSIONS:
* INITIAL MODELS
    + BMI as continuous
      - QUESTION: According to mean graph, linear relationship changes when BMI is under 30, 30-35, and over 35. How do I do an accurate linear regression? Should I subset the data and do three separate linear regressions for BMI under 30, 30-35, and over 40? 
    + Age as continuous
    + Age and BMI combined
      - QUESTION: Again, should I subset the data according to the changes in the linearity of BMI?
* ASSESSMENT: 
    + See which regression shows the highest R^2 

###LOGISTIC REGRESSIONS:   
* INITIAL MODEL includes all independent variables:
    - Smoker
    - Bmi (categorical)
    - Sex
    - Region
    - Children
    - Age
* ASSESSMENT: 
    - See which are strong predictors
    - Create new and improved model 
    - Calculate probabilities for high charges from each facet used in final model
    
###CLUSTERING:
* VARIABLES- all variables as binary (including charges). 
* PROCESS 1: 
    + Heigharachial dendrogram if R allows, to choose # of clusters
    + Apply clustering
* PRCESS 1 ASSESSMENT: Find percentage of "high" charges found in each group 
* (PROCESS 2): 
    + Learn how to do something meaningful using the last kmeans exercise I couldn't do...