---
title: "machine learning approach"
author: "Julia Sheriff"
date: "10/28/2018"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

###GENERAL IDEA OF DATASET

```{r}
hc1 <- read.csv("health_charges_clean.csv")
colnames(hc1)
hc1 <- hc1[ c(-1)]
str(hc1)
head(hc1)
hc2 <- read.csv("binary_charges.csv")
colnames(hc2)
hc2 <- hc2[ c(-1)]
str(hc2)
head(hc2)
```

###LINEAR REGRESSIONS:
* Checking class of variables: 
```{r}
str(hc1$smoker)
str(hc1$region)
str(hc1$sex)
str(hc1$children)
#because sex, region, and smoker are factors, the regression output will give a comparative interpretion of category against all the others in that facet.
#children is integer, so children will be treated as a variable in itself.
```

* Linear model:
* Adjusted R-squared is .7494, so the 75% of the value of the charges can be attributed to these variables.
* Significant variables:
    + being a smoker increases charges by $23848.9
    + having higher numbers of children increases charges by $475.5
    + higher bmi increases charges by $339.2
    + higher age increases charges by $256.9

```{r}
lmall <- lm ( charges ~ bmi + age + smoker + sex + children + region, data = hc1)
summary(lmall, method = lm)
plot(lmall)
```

###LOGISTIC REGRESSIONS:   
Create testing set:
```{r}
library(caTools)
set.seed(88)  

split  = sample.split(hc1$charges_factor, SplitRatio = .75 ) 
hc1train = subset(hc1, split == TRUE)
hc1test= subset(hc1, split == FALSE)
```
###Logistic regression:

* Important variables:
    + Being a smoker.
    + Age. Being in 40s was a less significant predictor than other age brackets. 
    + BMI: overweight, obese1, obese2
```{r}
lgall= glm(charges_factor ~ bmi_factor + age_factor + smoker + children + sex + region,  data = hc1train, family = binomial)
summary(lgall)
plot(lgall)
```

* Prediction:
    + 25% ability to prdict high charges, and 92% ability to predict low charges?
```{r}
predicttrain = predict(lgall, type = "response")
tapply(predicttrain, hc1train$charges_factor, mean) 
```

* Confusion matrix on training set with .5 threshhold:
    + 93.02% accuracy of predicting high health charges
```{r}
table(hc1train$charges_factor, predicttrain > .5)
(193 + 740) / ( 193 + 58 + 12 + 740)
```

* Confusion matrix on testing set with .5 threshhold:
    + 91.94% accuracy of predicting high health charges
```{r}
predicttest = predict(lgall, type = "response", newdata = hc1test)
table(hc1test$charges_factor, predicttest > .5)
(62 + 246) / ( 62 + 22 + 5 + 246)
```


###CLUSTERING:
* Heigharchial clustering with dendrogram of all variables aside from charge.
```{r}
colnames(hc2)
distances = dist(hc2[c(-1, -2)], method = "euclidian")
cluster1 = hclust(distances, method = "ward")
plot(cluster1)
```

*Eight clusters: groups with high charges:
    + Group 1 was a predictor of high charges at 91.7%
    + The percentage of high charges and the percent of smokers within each cluster were equal across all clusters. 
    + Smoking and high charges were most powerful in the clustering algorithm.
```{r}
clustergroups = cutree(cluster1, k = 8)
str(clustergroups)
highv <- tapply(hc2$charges_factor_high, clustergroups, mean)
highv <- as.vector(highv) 
smokev <- tapply(hc2$smoker_yes, clustergroups, mean)
smokev <- as.vector(highv)
nev <- tapply(hc2$region_northeast, clustergroups, mean)
nev <- as.vector(nev)
nwv <- tapply(hc2$region_northwest, clustergroups, mean)
nwv <- as.vector(nwv)
swv <- tapply(hc2$region_southwest, clustergroups, mean)
swv <- as.vector(swv)
sev <- tapply(hc2$region_southeast, clustergroups, mean)
sev <- as.vector(sev)
sexfv <- tapply(hc2$sex_female, clustergroups, mean)
sexfv <- as.vector(sexfv)
ch0v <- tapply(hc2$children_0, clustergroups, mean)
ch0v <- as.vector(ch0v)
ch1v <- tapply(hc2$children_1, clustergroups, mean)
ch1v <- as.vector(ch1v)
ch2v <- tapply(hc2$children_2, clustergroups, mean)
ch2v <- as.vector(ch2v)
ch3v <- tapply(hc2$children_3, clustergroups, mean)
ch3v <- as.vector(ch3v)
ch4v <- tapply(hc2$children_4, clustergroups, mean)
ch4v <- as.vector(ch4v)
ch5v <- tapply(hc2$children_5, clustergroups, mean)
ch5v <- as.vector(ch5v)
bmiuv <- tapply(hc2$bmi_factor_underweight, clustergroups, mean)
bmiuv <- as.vector(bmiuv)
bmihv <- tapply(hc2$bmi_factor_healthy_weight, clustergroups, mean)
bmihv <- as.vector(bmihv)
bmiov <- tapply(hc2$bmi_factor_overweight, clustergroups, mean)
bmiov <- as.vector(bmiov)
bmio1v <- tapply(hc2$bmi_factor_obese_1, clustergroups, mean)
bmio1v <- as.vector(bmio1v)
bmio2v <- tapply(hc2$bmi_factor_obese_2, clustergroups, mean)
bmio2v <- as.vector(bmio2v)
bmio3v <- tapply(hc2$bmi_factor_obese_3, clustergroups, mean)
bmio3v <- as.vector(bmio3v)

clusterframe <- cbind(highv, smokev, sexfv, bmiuv, bmihv, bmiov, bmio1v, bmio2v, bmio3v, ch0v, ch1v, ch2v, ch3v, ch4v, ch5v, nev, nwv, sev, swv )

View(clusterframe)
```

```
###CLUSTERING WITHOUT SMOKERS DATA
*more symmetrical dendrogram; high charges distributed relatively evenly between clusters.
```{r}
colnames(hc2)
hc3 <- hc2[ , c(1:14, 17:22)]
colnames(hc3)

distancesS = dist(hc3[c(-1, -2)], method = "euclidian")
clusterS = hclust(distancesS, method = "ward")
plot(clusterS)

```

```{r}
clustergroupsS = cutree(clusterS, k = 8)
str(clustergroupsS)
highvS <- tapply(hc2$charges_factor_high, clustergroupsS, mean)
highvS <- as.vector(highvS) 
smokevS <- tapply(hc2$smoker_yes, clustergroupsS, mean)
smokevS <- as.vector(highvS)
nevS <- tapply(hc2$region_northeast, clustergroupsS, mean)
nevS <- as.vector(nevS)
nwvS <- tapply(hc2$region_northwest, clustergroupsS, mean)
nwvS <- as.vector(nwvS)
swvS <- tapply(hc2$region_southwest, clustergroupsS, mean)
swvS <- as.vector(swvS)
sevS <- tapply(hc2$region_southeast, clustergroupsS, mean)
sevS <- as.vector(sevS)
sexfvS <- tapply(hc2$sex_female, clustergroupsS, mean)
sexfvS <- as.vector(sexfvS)
ch0vS <- tapply(hc2$children_0, clustergroupsS, mean)
ch0vS <- as.vector(ch0vS)
ch1vS <- tapply(hc2$children_1, clustergroupsS, mean)
ch1vS <- as.vector(ch1vS)
ch2vS <- tapply(hc2$children_2, clustergroupsS, mean)
ch2vS <- as.vector(ch2vS)
ch3vS <- tapply(hc2$children_3, clustergroupsS, mean)
ch3vS <- as.vector(ch3vS)
ch4vS <- tapply(hc2$children_4, clustergroupsS, mean)
ch4vS <- as.vector(ch4vS)
ch5vS <- tapply(hc2$children_5, clustergroupsS, mean)
ch5vS <- as.vector(ch5vS)
bmiuvS <- tapply(hc2$bmi_factor_underweight, clustergroupsS, mean)
bmiuvS <- as.vector(bmiuvS)
bmihvS <- tapply(hc2$bmi_factor_healthy_weight, clustergroupsS, mean)
bmihvS <- as.vector(bmihvS)
bmiovS <- tapply(hc2$bmi_factor_overweight, clustergroupsS, mean)
bmiovS <- as.vector(bmiovS)
bmio1vS <- tapply(hc2$bmi_factor_obese_1, clustergroupsS, mean)
bmio1vS <- as.vector(bmio1vS)
bmio2vS <- tapply(hc2$bmi_factor_obese_2, clustergroupsS, mean)
bmio2vS <- as.vector(bmio2vS)
bmio3vS <- tapply(hc2$bmi_factor_obese_3, clustergroupsS, mean)
bmio3vS <- as.vector(bmio3vS)

clusterframeS <- cbind(highvS, smokevS, sexfvS, bmiuvS, bmihvS, bmiovS, bmio1vS, bmio2vS, bmio3vS, ch0vS, ch1vS, ch2vS, ch3vS, ch4vS, ch5vS, nevS, nwvS, sevS, swvS )

View(clusterframeS)
```